MYC uncouples HIF target gene expression from glycolytic flux in hypoxic proliferating primary cells

### 2021-12-07 manuscript/manuscript.Rmd
update manuscript with MYC data

---
title: "Hypoxia uncouples HIF target gene transcription and glycolytic flux in proliferating primary cells"

@@

bibliography: 
- library.bib
- packages.bib
csl: cell-metabolism.csl
link-citations: yes

output: 
  bookdown::word_document2:
    reference_docx: !expr system.file("manuscript", "template.docx", package = "Copeland.2021.hypoxia.flux")
    df_print: kable
    fig_caption: true
    number_sections: false
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
      - '--lua-filter=pagebreak.lua'
      # - '--lua-filter=multiple-bibliographies.lua'

compact-title: true
geometry: margin = 0.75in
fontsize: 11pt
mainfont: Calibri
graphics: yes
---

@@

biomass <- targets::tar_read(biomass)
resources <- targets::tar_read(resources_table)
biomass <- tibble::deframe(biomass)
m5 <- targets::tar_read(pasmc_m5)
targets::tar_load(
  c(
    m1_figure, 
    m2_figure, 
    m3_figure,
    m4_figure, 
    m5_figure, 
    m6_figure
  )
)
```

@@

# Summary 
Hypoxia is an important environmental stressor that causes transcriptional and metabolic reprogramming in cells to facilitate their survival. Here, we performed stable isotope tracing and metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia. Despite activation of the hypoxia-inducible factor (HIF) transcriptional program and increased expression of glycolytic enzymes, glycolytic flux was decreased in hypoxic cells. While pharmacologic stabilization of HIF in normoxia with the prolyl hydroxylase inhibitor molidustat did increase glycolytic flux as expected, hypoxia abrogated this effect of molidustat treatment. Multi-omic profiling of cells treated with hypoxia or molidustat, separately or together, revealed distinct molecular responses to hypoxia and pharmacologic prolyl hydroxylase inhibition. Together, these data suggest that primary cell bioenergetic metabolism is closely coupled to cell proliferation rate, and that other factors supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux.

@@

hypoxia, metabolic flux analysis, hypoxia-inducible factor, prolyl hydroxylase, metabolism

@@

Cellular responses to ambient oxygen levels drive numerous physiologic and pathologic processes from wound healing and angiogenesis to pulmonary vascular remodeling and fibrosis [@RN1980]. These responses require a sustained capacity for cell proliferation, migration, and protein synthesis, with their associated energetic and metabolic demands, even in the face of limited oxygen availability, or hypoxia. Metazoan cells depend on aerobic respiration to meet cellular energy needs. With an inadequate oxygen supply, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation. Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of enzymes in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980; @RN1198]. Although HIF-1α is constitutively expressed, it is hydroxylated by prolyl hydroxylase enzymes (PHDs) in normoxia and targeted for proteasomal degradation. PHDs are the principal oxygen sensors in metazoan cells [@RN999]. PHDs are α-ketoglutrate-dependent dioxygenase enzymes that require molecular oxygen for their enzymatic activity. When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and activation of its associated transcriptional program. Overall, the HIF-1 transriptional program should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and to minimize the formation of reactive oxygen species (ROS) [@RN2525].

@@

To address these questions, here we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen. These cells were selected as they may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of human lung diseases where tissue hypoxia is a prominent feature. We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1 transcriptional program. In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia abrogated this response. These findings suggest the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1-dependent up-regulation of glycolytic gene expression in human primary cells.

@@

The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary LFs and PASMCs. Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:m1)A**), thought to be an indicator of metabolic steady state . LFs cultured in 0.5% oxygen grew slower than LFs cultured in 21% oxygen (**Figure \@ref(fig:m1)B**), but slower growth was not associated with decreased cell viability (**Figure S1A**). These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:m1)K, E-H**). These changes persisted for the duration of the experimental time course.

@@

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figures \@ref(fig:m1)I-J**). Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Figure S1**). Surprisingly, we observed no significant differences in glucose uptake or lactate efflux rates in 0.5% oxygen cultures (**Figure \@ref(fig:m1)I**) despite activation of the HIF-1 transcriptional program as reflected by increased expression of GLUT1 and LDHA. No change in glucose or lactate fluxes were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**). Similar results were also observed in PASMCs were lactate efflux was significantly *decreased* by 0.5% ambient oxygen (**Figure S3**).

In addition to glucose and lactate, we also determinted the extracellular fluxes of pyruvate and amino acids. Overall, changes were modest, with hypoxia generally associated with decreased fluxes of all measured metabolites. Notably, a substantial decrease in glutamine consumption was observed in LFs cultured in 0.5% oxygen. This observation is in contrast to previous studies in cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia in these cells [@RN1447; @RN2395; @RN628]. Similar patterns were observed under 0.2% oxygen culture conditions. In PASMCs, glutamine uptake did increase, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), highlighting differential responses of these mesenchymal cells to hypoxia.

Given that hypoxia did not increase glucose and lactate fluxes as expected from the associated increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs. Cells were treated with the prolyl hydroxylase inhibitor BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:m1)**). Similar to hypoxia, BAY decreased cell growth rate (**Figures \@ref(fig:m1)A-B**) and activated the HIF-1 transcriptional program (**Figures \@ref(fig:m1)D-H**). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:m1)I**). Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:m1)J**) with preservation of glutamine uptake, alanine efflux, and glutamate efflux rates.

@@

To validate our findings from extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of label into key carbon utilization pathways (**Figures \@ref(fig:m2), S4-S5**). Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:m3)A-B**).

@@

The stable isotope labeling patterns in PASMCs were similar to LFs (**Figure S5**) where labeling was attenuated in hypoxia. The most notable differences between LF and PASMC labeling were observed in citrate. Compared with LFs, PASMCs demonstrated a marked reduction of M2-citrate labeled by glucose and no change in the labeling of M3-citrate by [U-^13^C~3~] glucose. Interestingly, unlike LFs, PASMCs demonstrated increased production of M5-citrate from [U-^13^C~5~] glutamine in hypoxia (`r m5[["mean_21%"]]`% ± `r m5[["se_21%"]]`% *v*. `r m5[["mean_0.5%"]]`% ± `r m5[["se_0.5%"]]`%, p = `r m5[["pval"]]`) consistent with increased reductive carboxylation in support of lipid synthesis in these cells, similar to prior observations in cancer cells.

@@

The mass isotopomer distribution for a given metabolite is determined based on a complicated relationship among the rate of isotope incorporation into the metabolic network, the contributions of unlabeled substrates, and fluxes through related pathways. To clarify how these labeling patterns reflect changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above. Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figures \@ref(fig:m3)A, S6**). Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:m3), S7, Tables S1-S3**). Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis. Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Figure S7A**). Consistent with extracellular flux measurements and isotope labeling patterns described above, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed in the metabolic flux models of LFs cultured in hypoxia. A significant increase in pentose phosphate pathway flux was also observed, although the absolute flux through this pathway is low (**Figure \@ref(fig:m3)A**). By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:m3)B**), but had a similar effect as hypoxia in decreasing serine and glutamine incorporation. Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.

@@

Second, PHD inhibition is associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,500 and 700 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively, with similar results in PASMCs. Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment. This finding may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366]. To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figures \@ref(fig:m4), S5**). Notably, lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high. Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@RN2600] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*. In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F ]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization. Subsequently, investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718]. With hypoxia or BAY treatment, we observed increased labeling of the TCA metabolites citrate (CIT), α-ketoglutarate (AKG), malate (MAL), and aspartate (ASP) in LFs. Interestingly, although increased labeling of pyruvate and lactate were observed in hypoxic PASMCs, the label was not incorporated into the TCA cycle as observed in LFs (**Figure S5**).

@@

Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen (**Figure \@ref(fig:m5)D**). Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells. These observations are again consistent with the results of metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition. To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different than the change with BAY treatment). Of 133 metabolites, 77 were significantly differentially regulated by hypoxia and BAY treatments. Several patterns emerged from this analysis (**Figure \@ref(fig:m5)F**). Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate [@RN1603], increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs. The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine. The glycolytic intermediate glyceraldehyde 3-phosphate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA). A metabolite set enrichment analysis of these differentially regulated metabolites revealed the TCA cycle to be the most enriched KEGG metabolite set (**Figure \@ref(fig:m5)G**). Leading edge analysis demonstrated negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Figure \@ref(fig:m5)H**). This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:m3)**).

@@

To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Figure \@ref(fig:m6), S9**). As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Figure S9A, S9B**). Of the 7789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1649 (21%) were shared between BAY and hypoxia, while 5249 (67%) were unique to 0.5% culture (**Figure S9C**). Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@RN3112] (**Figures S9D, S9E, S9F**). As expected, both treatments were associated with enrichment of the "hypoxia" and "glycolysis" gene sets.

@@

Interestingly, an enrichment analysis of transcripts differentially regulated by 0.5% oxygen and BAY treatment actually demonstrated greater enrichment of the "hypoxia" gene set following BAY treatment alone than with hypoxia itself (**Figure \@ref(fig:m6)D**). Similarly, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment. The most enriched gene sets associated with hypoxia included pro-proliferative "E2F targets" and "G2/M checkpoint" proteins as well as "Myc targets". These findings were further supported by a transcription factor enrichment analysis identifying enrichment of Myc target proteins associated with hypoxia (**Figure \@ref(fig:m6)E**). Consistent with these bioinformatic results, immunoblotting demonstrated up-regulation of Myc protein in hypoxia-treated cells, but not with BAY-treatment alone (**Figure \@ref(fig:m6)F**). These data provide a starting point for interrogating the relative hierarchy of transcription factor regulation of gene expression by hypoxia in proliferating primary cells and suggest several potential mechanisms that may contribute to the uncoupling of glycolytic gene transcription and glycolytic flux reported here.

@@

>>>>

---
title: "MYC uncouples HIF target gene expression from glycolytic flux in hypoxic proliferating primary cells"

@@

bibliography: 
- library.bib
- packages.bib
csl: cell-metabolism.csl
link-citations: yes
compact-title: true
geometry: margin = 0.75in
fontsize: 11pt
mainfont: Calibri
graphics: yes
---

@@

biomass <- targets::tar_read(biomass)
resources <- targets::tar_read(resources_table)
biomass <- tibble::deframe(biomass)
m5 <- targets::tar_read(pasmc_m5)
targets::tar_load(
  c(
    m1_figure, 
    m2_figure, 
    m3_figure,
    m4_figure, 
    m5_figure, 
    m6_figure, 
    m7_figure
  )
)
```

@@

# Summary

Hypoxia requires metabolic adaptations to sustain cellular functions that underlie numerous physiologic and pathologic processes. While many studies have explored the metabolic consequences of hypoxia in cancer models, comparatively little is known about the metabolic response of primary cells to hypoxia. Here, we performed metabolic flux analyses of proliferating human lung fibroblasts and pulmonary artery smooth muscle cells in hypoxia. Unexpectedly, glycolytic flux was decreased in hypoxic cells despite activation of the hypoxia-inducible factor (HIF) transcriptional program and increased expression of glycolytic enzymes. Pharmacologic activation of HIF with the prolyl hydroxylase (PHD) inhibitor molidustat in normoxia did increase glycolytic flux, but hypoxia abrogated this effect. Multi-omic profiling of cells treated with hypoxia or molidustat, separately or together, revealed distinct molecular responses to hypoxia and pharmacologic PHD inhibition and suggested a critical role for MYC in modulating the HIF response in hypoxia. MYC knockdown in hypoxia increased lactate efflux while MYC overexpression in normoxia blunted the effects of molidustat treatment. Together, these data suggest that other factors, including MYC, supersede the anticipated effects of HIF-dependent up-regulation of glycolytic gene expression on glycolytic flux in hypoxic proliferating primary cells.

@@

hypoxia, metabolic flux analysis, hypoxia-inducible factor, prolyl hydroxylase, metabolism, MYC, pulmonary artery smooth muscle cell, lung fibroblast

@@

Cellular responses to ambient oxygen levels drive numerous physiologic and pathologic processes from wound healing and angiogenesis to pulmonary vascular remodeling and fibrosis [@RN1980]. These responses require a sustained capacity for cell proliferation, migration, and protein synthesis, with their associated energetic and metabolic demands, even in the face of limited oxygen availability, or hypoxia. Metazoan cells depend on aerobic respiration to meet cellular energy needs. With an inadequate oxygen supply, cells must reduce energy consumption and shift energy production away from oxidative phosphorylation. Cells accomplish this goal through stabilization of the hypoxia-inducible transcription factor 1α (HIF-1α), which activates the transcription of glucose transporters, glycolytic enzymes, lactate dehydrogenase, and pyruvate dehydrogenase kinase, while decreasing the expression of enzymes in the tricarboxylic acid (TCA) cycle and electron transport chain [@RN1980; @RN1198]. Although HIF-1α is constitutively expressed, it is hydroxylated by prolyl hydroxylase enzymes (PHDs) in normoxia and targeted for proteasomal degradation. PHDs are the principal oxygen sensors in metazoan cells [@RN999]. PHDs are α-ketoglutarate-dependent dioxygenase enzymes that require molecular oxygen for their enzymatic activity. When oxygen tension falls, PHD activity decreases, leading to HIF-1α stabilization and activation of its associated transcriptional program. Overall, the HIF-1 transcriptional program should increase glycolytic capacity and divert glucose-derived pyruvate from oxidative phosphorylation toward lactate fermentation to maintain ATP production and to minimize the formation of reactive oxygen species (ROS) [@RN2525].

@@

To address these questions, we have developed models of bioenergetic carbon flux in human lung fibroblasts (LFs) and pulmonary artery smooth muscle cells (PASMCs) cultured in 21% or 0.5% oxygen. These cells were selected as they may be exposed to a wide range of oxygen concentrations *in vivo*, continue to proliferate despite hypoxic culture conditions *in vitro*, and play important roles in the pathology of human lung diseases where tissue hypoxia is a prominent feature. We found that hypoxia fails to increase glycolysis in these primary cells despite robust up-regulation of the HIF-1 transcriptional program. In normoxia, HIF-1α stabilization by the PHD inhibitor molidustat (BAY-85-3934, "BAY") [@RN580] did increase glycolysis and lactate efflux; however, hypoxia blocked this response. These findings suggested the existence of important hypoxia-dependent regulatory mechanisms that override the metabolic consequences of HIF-1-dependent up-regulation of glycolytic gene expression in human primary cells. Transcriptomic profiling suggested a critical role for the transcription factor c-MYC (MYC) in the adaptive response to hypoxia. Using knockdown and overexpression approaches, we demonstrate that MYC attenuates HIF-driven glycolysis in hypoxia and following BAY treatment.

@@

The goal of this study was to identify the metabolic changes associated with hypoxia in proliferating primary LFs and PASMCs. Cells were seeded and placed into hypoxia for 24 h prior to sample collection to provide adequate time for activation of the hypoxia-dependent transcriptional program. From this starting point, we identified the optimal cell seeding density and time course to capture exponential cell growth (**Figure \@ref(fig:m1)A**), thought to be an indicator of metabolic steady state. LFs cultured in 0.5% oxygen grew slower than LFs cultured in 21% oxygen (**Figure \@ref(fig:m1)B**), but slower growth was not associated with decreased cell viability (**Figure S1A**). These cells demonstrated robust stabilization of HIF-1α protein associated with up-regulation of downstream targets, such as glucose transporter 1 (GLUT1) and lactate dehydrogenase A (LDHA) (**Figures \@ref(fig:m1)K, E-H**). These changes persisted for the duration of the experimental time course.

@@

Having established a model system, we next determined the extracellular fluxes of glucose (GLC), lactate (LAC), pyruvate (PYR), and amino acids (**Figures \@ref(fig:m1)I-J**). Flux calculations incorporated changes in cell number, extracellular metabolite concentrations, and medium evaporation over time [@RN1523] (**Figure S1**). Surprisingly, we observed no significant differences in glucose uptake or lactate efflux rates in 0.5% oxygen cultures (**Figure \@ref(fig:m1)I**) despite activation of the HIF-1 transcriptional program as reflected by increased expression of GLUT1 and LDHA. No change in glucose or lactate fluxes were observed when the ambient oxygen level was decreased further to 0.2% (**Figure S2**). Similar results were also observed in PASMCs where lactate efflux was significantly *decreased* by 0.5% ambient oxygen (**Figure S3**).

In addition to glucose and lactate, we also determined the extracellular fluxes of pyruvate and amino acids. Overall, changes were modest, with hypoxia generally decreasing the fluxes of all measured metabolites. Notably, a substantial decrease in glutamine consumption was observed in LFs cultured in 0.5% oxygen. This observation is in contrast to previous studies in cancer cell metabolism demonstrating increased glutamine uptake as a key feature of the metabolic response to hypoxia [@RN1447; @RN2395; @RN628]. Similar patterns were observed in 0.2% oxygen (**Figure S2**). In PASMCs, glutamine uptake did increase, as did the uptake of branched-chain amino acids and arginine (**Figure S3**), highlighting differential responses of these mesenchymal cells to hypoxia.

Given that hypoxia did not increase glucose and lactate fluxes as expected based on the observed increases in glycolytic gene expression, we next assessed the capacity of HIF-1α to augment glycolysis in LFs. Cells were treated with the prolyl hydroxylase inhibitor BAY to stabilize HIF-1α under 21% oxygen conditions (**Figure \@ref(fig:m1)**). Similar to hypoxia, BAY decreased cell growth rate (**Figures \@ref(fig:m1)A-B**) and activated the HIF-1 transcriptional program (**Figures \@ref(fig:m1)D-H**). Compared to hypoxia, BAY treatment resulted in a similar activation of HIF-1 target gene transcription and protein expression. In normoxia, this transcriptional program was associated with substantially increased glucose uptake and lactate efflux (**Figure \@ref(fig:m1)I**). Comparatively modest effects of BAY on amino acid fluxes were observed as compared to 0.5% oxygen culture conditions (**Figure \@ref(fig:m1)J**) with preservation of glutamine uptake, alanine efflux, and glutamate efflux rates.

@@

To validate the findings from our extracellular flux measurements, we next treated LFs with stable carbon isotopes of glucose and glutamine to measure the rate and trace the incorporation of ^13^C into key downstream carbon utilization pathways (**Figures \@ref(fig:m2), S4-S5**). Consistent with our extracellular flux measurements, we found that the rate of label incorporation from [U-^13^C~6~] glucose into pyruvate was similar across 21%, 0.5%, and DMSO-treated conditions, but markedly increased with BAY treatment (**Figures \@ref(fig:m3)A-B**).

@@

The stable isotope labeling patterns in PASMCs were similar to LFs (**Figure S5**) where labeling was attenuated in hypoxia. The most notable differences between LF and PASMC labeling were observed in citrate. Compared with LFs, PASMCs demonstrated a marked reduction of M2-citrate labeled by glucose and no change in the labeling of M3-citrate by [U-^13^C~6~] glucose. These data suggest decreased glucose entry into the TCA cycle *via* pyruvate dehydrogenase and no change in the activity of pyruvate carboxylase, respectively. Interestingly, unlike LFs, PASMCs demonstrated increased production of M5-citrate from [U-^13^C~5~] glutamine in hypoxia (`r m5[["mean_21%"]]`% ± `r m5[["se_21%"]]`% *v*. `r m5[["mean_0.5%"]]`% ± `r m5[["se_0.5%"]]`%, p = `r m5[["pval"]]`) consistent with increased reductive carboxylation in support of lipid synthesis in these cells, similar to prior observations in cancer cells, but distinct from LFs studied here.

@@

The mass isotopomer distribution for a given metabolite is determined based on a complicated relationship among the rate of isotope incorporation into the metabolic network, the contributions of unlabeled substrates, and fluxes through related pathways. To clarify how these labeling patterns reflect changes in intracellular metabolite fluxes, we next generated metabolic flux models incorporating the extracellular flux measurements and stable isotope tracing data described above. Preliminary labeling time courses indicated that, even after 72 h of labeling, intracellular metabolites did not reach isotopic steady state (**Figure S6**). Thus, we performed isotopically non-stationary metabolic flux analysis as implemented by INCA [@RN1522; @RN2501; @RN964] (**Figures \@ref(fig:m3), S7, Tables S1-S3**). Overall, proliferating LFs demonstrated high rates of glucose uptake and glycolysis. Approximately 10% of cytoplasmic pyruvate enters the TCA cycle with the balance converted to lactate (**Figure S7A**). Consistent with extracellular flux measurements and isotope labeling patterns described above, significant reductions in glycolysis, the TCA cycle, and amino acid metabolism were observed in the metabolic flux models of LFs cultured in hypoxia. A significant increase in pentose phosphate pathway flux was also observed, although the absolute flux through this pathway is low (**Figure \@ref(fig:m3)A**). By contrast, HIF-1 activation by BAY in 21% oxygen increased glycolysis and lactate fermentation by ~50% (**Figure \@ref(fig:m3)B**), but had a similar effect to hypoxia in decreasing serine and glutamine incorporation. Metabolite fluxes in DMSO-treated cells were similar to 21% oxygen controls.

@@

Second, PHD inhibition is associated with a marked increase in the lactate transport exchange flux in LFs from ~ 0 to 1,500 and 700 fmol/cell/h in 0.5% oxygen and BAY treatment conditions, respectively, with similar results in PASMCs. Since the net lactate transport flux is secretion, this observation suggests increased lactate uptake with hypoxia or BAY treatment. This finding may be consistent with the HIF-driven increased expression of the reversible lactate transporter MCT4 [@RN366]. To investigate this hypothesis, LFs and PASMCs were treated with [U-^13^C~3~] lactate (2 mM) and ^13^C incorporation into intracellular metabolites was analyzed by LC-MS (**Figures \@ref(fig:m4), S5**). Notably, lactate labeled ~50% of citrate and ~20% of downstream TCA cycle metabolites in both LFs and PASMCs, indicating that lactate may be an important respiratory fuel source in these cells even though lactate efflux is high. Although lactate has been used less commonly than glucose and glutamine in stable isotope tracing studies, Faubert and colleagues [-@RN2600] demonstrated lactate incorporation in human lung adenocarcinoma *in vivo*. In this study, lactate incorporation corresponded to regions of high glucose uptake as determined by [ ^18^F ]-fluorodeoxyglucose positron emission tomography, suggesting that lactate consumption can occur in areas of high glucose utilization. Subsequently, investigators have demonstrated the importance of lactate as a metabolic fuel *in vivo* [@RN2756; @RN2718]. With hypoxia or BAY treatment, we observed increased labeling of the TCA metabolites citrate (CIT), α-ketoglutarate (AKG), malate (MAL), and aspartate (ASP) in LFs. Interestingly, although increased labeling of pyruvate was observed in hypoxic PASMCs, the label was not incorporated into the TCA cycle as observed in LFs (**Figure S5**).

@@

Principal component analysis revealed greater class similarity among both treatment groups cultured in 0.5% oxygen than among the BAY-treatment groups (**Figure \@ref(fig:m5)D**). Moreover, these hypoxia-treated cells were well-segregated from BAY-treated cells. These observations are again consistent with the results of the metabolic flux models demonstrating an overriding effect of hypoxia *per se* on metabolic flux and highlighting important differences between hypoxic and pharmacologic PHD inhibition. To identify the key differences among these treatments, we performed linear modeling focusing on those metabolites with a significant interaction term in the model described by treatment × oxygen (**Figure \@ref(fig:m5)E**) (*i.e.*, metabolites where the change with hypoxia treatment was different than the change with BAY treatment). Of 133 metabolites, 77 were significantly differentially regulated by hypoxia and BAY treatments. Several patterns emerged from this analysis (**Figure \@ref(fig:m5)F**). Interestingly, the hypoxia-regulated metabolite, 2-hydroxyglutarate [@RN1603], increased following BAY treatment, but was unaffected by hypoxia treatment in proliferating LFs. The tricarboxylic acid (TCA) cycle metabolite aconitate was decreased in hypoxia, but increased by BAY treatment while the opposite effect was observed for hydroxyproline and taurine. The glycolytic intermediate glyceraldehyde 3-phosphate (GAP) was decreased by both hypoxia and BAY, while hypoxia reversed the BAY-mediated decrease in γ-aminobutyric acid (GABA). A metabolite set enrichment analysis of these differentially regulated metabolites revealed the TCA cycle to be the most enriched KEGG metabolite set (**Figure \@ref(fig:m5)G**). Leading edge analysis demonstrated negative enrichment scores associated with all of the TCA metabolites detected by our platform (**Figure \@ref(fig:m5)H**). This result suggests a more modest impact of BAY treatment on the TCA cycle than hypoxia, as suggested by our metabolic flux models where hypoxia resulted in a 1.5-2-fold reduction of TCA flux compared to a 1.1-1.5-fold reduction with BAY treatment in normoxia (**Figure \@ref(fig:m3)**).

@@

To identify the upstream regulators of the observed metabolic changes, we next performed transcriptomic analysis of LFs treated with hypoxia or BAY, separately or together (**Figure \@ref(fig:m6), S9**). As anticipated, both hypoxia and BAY treatment induced substantial changes in gene expression (**Figure S9A, S9B**). Of the 7,789 differentially expressed genes across both conditions, 891 (11%) were unique to BAY treatment in normoxia, 1,649 (21%) were shared between BAY and hypoxia, while 5,249 (67%) were unique to 0.5% culture (**Figure S9C**). Gene set enrichment analysis of these differentially regulated metabolites was performed using Molecular Signatures Database "Hallmark" gene sets [@RN3112] (**Figures S9D, S9E, S9F**). As expected, both treatments were associated with enrichment of the "hypoxia" and "glycolysis" gene sets.

@@

Interestingly, an enrichment analysis of transcripts differentially regulated by 0.5% oxygen and BAY treatment actually demonstrated greater enrichment of the "hypoxia" gene set following BAY treatment alone than with hypoxia itself (**Figure \@ref(fig:m6)D**). Similarly, the gene set for glycolysis was also more enriched following BAY treatment than hypoxia treatment. The most enriched gene sets associated with hypoxia included pro-proliferative "E2F targets" and "G2/M checkpoint" proteins as well as "MYC targets". These findings were further supported by a transcription factor enrichment analysis identifying enrichment of MYC target proteins associated with hypoxia (**Figure \@ref(fig:m6)E**).

## MYC antagonizes HIF-dependent glycolytic fluxes

Consistent with our bioinformatic results, immunoblotting demonstrated increased MYC protein levels in hypoxia-treated cells, but not with BAY-treatment alone, where MYC was decreased (**Figure \@ref(fig:m7)A-B**). To test the hypothesis that hypoxia-induced MYC expression inhibits glycolysis in hypoxic primary cells, we first combined MYC knockdown with hypoxia treatment (**Figure \@ref(fig:m7)C-E**). As expected, MYC deficient cells proliferated more slowly in normoxia and MYC was absolutely essential to sustain cell proliferation in hypoxia (**Figure \@ref(fig:m7)D**). Consistent with our hypothesis, MYC knockdown cells demonstrated increased lactate efflux upon hypoxia treatment, unlike control siRNA-treated cells (**Figure \@ref(fig:m7)E**). We next performed the complementary experiment to determine whether MYC overexpression could attenuate the increase in glycolysis observed with BAY treatment (**Figure \@ref(fig:m7)F-H**). MYC increased the proliferation rate of DMSO-treated cells, although did not augment the proliferation rate of BAY-treated cells. As expected, MYC overexpression blocked the BAY-stimulated increase in lactate efflux. Together, these data suggest that hypoxia-induced MYC expression may be one factor that uncouples the HIF transcriptional program from glycolytic flux in proliferating primary cells.

@@

<!-- MYC role in metabolism -->

Our data indicate that hypoxia-induced MYC expression is one such regulatory mechanism. MYC is a transcription factor that regulates the expression of numerous genes involved in many biological processes, including metabolism, proliferation, apoptosis, and differentiation [@dang2006; @li2020d; @stine2015]. As deregulated MYC activity has been associated with the majority of human cancers [@vita2006], much of our understanding of MYC regulation comes from studies using cancer cell models [@li2020d; @stine2015; @dang2012a; @madden2021], while the role of MYC in the biology of transformed cells is less well understood. The literature describes a complex and reciprocal relationship between HIF and MYC that depends on both environmental (*e.g.*, hypoxia) and cellular context [@li2020d]. Generally, HIF-1 has been observed to inhibit MYC through multiple mechanisms [koshiji2004; koshiji2005; @gordan2007; @zhang2007], and this previous work is consistent with our observations that HIF stabilization following BAY treatment decreased MYC protein and target gene expression. Conversely, MYC has been implicated in increased HIF activity through transcriptional and post-transcriptional mechanisms [@zhang2009a; @chen2013; @doe2012], primarily in the context of malignant transformation. The observation that MYC may antagonize the transcriptional effects of HIF to sustain primarily cell proliferation and metabolism in hypoxia suggests a substantially different regulatory relationship than has been previously described.
